---
title: "Lab1"
author: "Anastasia Piadi, Nazli Bilgic"
date: "`r Sys.Date()`"
output: pdf_document
header-includes:
  - "\\usepackage{newunicodechar}"
  - "\\newunicodechar{−}{-}"
---
Question 1:

a) Plot the log likelihood function for the given data in the range from -4 to 4. Plot the derivative in the same range and check visually how often the derivative is equal to 0.

From the derivative plot we can see that the derivative is equal to 0, 3 times.
```{r}
log_likeli <- function() {
  x <- c(-2.8, 3.4, 1.2, -0.3, -2.6)
  n <- length(x)
  theta_values <- seq(-4, 4, by = 0.01)

  log_likelihood <- sapply(theta_values, function(b) -n * log(pi) - sum(log(1 + (x - b)^2)))
  derivative <- sapply(theta_values, function(b) sum(2 * (x - b) / (1 + (x - b)^2)))

  # Plot log-likelihood
  plot(theta_values, log_likelihood, type = "l", col = "blue",
       main = "Log Likelihood Function",
       xlab = expression(theta), ylab = "Log Likelihood")

  # Plot the derivative
  plot(theta_values, derivative, type = "l", col = "red",
       main = "Derivative of the Log Likelihood Function",
       xlab = expression(theta), ylab = "Derivative")
  abline(h = 0, col = "purple", lty = 2)

}

log_likeli()

```
b) Choose one of the following methods: bisection, secant, or Newton-Raphson. Write your own code to optimize with the chosen method. If you have chosen Newton-Raphson, describe how you derived the second derivative.

source:https://en.wikipedia.org/wiki/Secant_method

```{r}
log_likeli <- function() {
  x <- c(-2.8, 3.4, 1.2,-0.3,-2.6)
  n <- length(x)
  theta_values <- seq(-4, 4, by = 0.01)

  derivative_func <- function(theta) {
    sum(2 * (x - theta) / (1 + (x - theta) ^ 2))
  }
  secant_method <-
    function(f,
             x0,
             x1,
             epsilon = 1e-8,
             max_iter = 100) {
      for (i in 1:max_iter) {
        fx1 <- f(x1)
        fx0 <- f(x0)
        if (fx1 - fx0 == 0) {
          stop("Division by zero in secant method.")
        }
        x2 <- x1 - fx1 * (x1 - x0) / (fx1 - fx0)
        if (abs(x2 - x1) < epsilon) {
          cat("Secant method converged at iteration:", i, "\n")
          return(x2)
        }
        x0 <- x1
        x1 <- x2
      }
      stop("Secant method did not converge within the specified number of iterations.")
    }

  # Find the root using the secant method with initial guesses
  root <- secant_method(derivative_func,-4, 4)


  return(root)
}
root <- log_likeli()
print(paste("Root:", root))

```

c) Choose suitable starting values based on your plots to identify all local maxima of the likelihood function. Note that you need pairs of interval boundaries for the bisection or pairs of starting values for secant. Are there any (pairs of) starting values which do not lead to a local maximum? Decide which is the global maximum based on programming results.





d)


Question 2:

a) Write your own R function, myvar, to estimate the variance in this way.

```{r Question2 a,echo=FALSE, fig.cap=""}
myvar <- function(x) {
  n <- length(x)
  return((sum(x^2) - (sum(x)^2) / n) / (n - 1))
}
ex_data<-c(1,2,3,4)
cat("myvar function with the example data(c(1,2,3,4)):",myvar(ex_data), "\n")
cat("var function with the example data(c(1,2,3,4)):",var(ex_data))
```

b) Generate a vector x = (x1, . . . , x10000) with 10000 random numbers with mean 108 and variance 1.


```{r Question2 b,echo=FALSE, fig.cap=""}
myvar <- function(x) {
  n <- length(x)
  return((sum(x^2) - (sum(x)^2) / n) / (n - 1))
}
set.seed(123)
n <- 10000
mean_value <- 10^8
variance<-1
std_dev <- sqrt(variance)
x <- rnorm(n, mean = mean_value, sd = std_dev)

cat("myvar function output with random data:",myvar(x), "\n")
cat("var function output with random data:",var(x))
```

c) For each subset Xi = {x1,...,xi}, i = 1,...,10000 compute the difference Yi = myvar(Xi) − var(Xi), where var(Xi) is the standard variance estimation function in R. Plot the dependence Yi on i. Draw conclusions from this plot. How well does your function work? Can you explain the behaviour?

The difference between myvar and var function can be attributed to the way floating point arithmetic works in computers. This can lead to precision errors, especially when dealing with a combination of very large and very small numbers.

source:https://en.wikipedia.org/wiki/Floating-point_arithmetic#IEEE_754:_floating_point_in_modern_computers

```{r Question2 c, echo=FALSE, fig.cap=""}
myvar <- function(x) {
  n <- length(x)
  return((sum(x^2) - (sum(x)^2) / n) / (n - 1))
}

set.seed(123)
n <- 10000
mean_value <- 10^8
variance<-1
std_dev <- sqrt(variance)
x <- rnorm(n, mean = mean_value, sd = std_dev)

computed_variance <- myvar(x)
difference_fnctns <- numeric(n)
for (i in 1:n) {
  x_sub <- x[1:i]
  difference_fnctns[i] <- myvar(x_sub) - var(x_sub)
}
plot(difference_fnctns)

```


d) How can you better implement a variance estimator? Find and implement a formula that will give the same results as var().

We used two-pass method to calculate variance in another way. In two-pass method, we first compute the sample mean then the sum of the squares of the differences from the mean. 

source: https://en.wikipedia.org/wiki/Algorithms_for_calculating_variance

```{r Question2 d,echo=FALSE, fig.cap=""}
library(ggplot2)

var_two_pass <- function(x) {
  n <- length(x) 
  mean_val <- sum(x) / n #mean calculation
  sum_sq_diffs <- sum((x - mean_val)^2) #sum squared differences from the mean
  return(sum_sq_diffs / (n - 1)) #variance calculate
}

set.seed(123)
n <- 10000
mean_value <- 10^8
variance <- 1
std_dev <- sqrt(variance)

num_datasets <- 100
differences <- numeric(num_datasets)

for (i in 1:num_datasets) {
  x <- rnorm(n, mean = mean_value, sd = std_dev)
  differences[i] <- var_two_pass(x) - var(x)
}

df <- data.frame(dataset = 1:num_datasets, difference = differences)
ggplot(df, aes(x = dataset, y = difference)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Difference between two-pass and var functions",
       x = "Dataset number",
       y = "Difference in variance")


```



#Appendix

##Appendix A: Code for the questions


```{r ref-appendix, echo=FALSE, results='asis'}
all_labels <- knitr::all_labels()
for (label in all_labels) {
  if (label != "ref-appendix") {  # Skip the appendix chunk itself
    cat("### Chunk Label: ", label, "\n\n")
    cat("```r\n")
    code <- knitr::knit_code$get(label)
    cat(paste(code, collapse = "\n"))  # Ensure each line is separated
    cat("\n```\n\n")
  }
}




